Almost orthogonal vectors

### 2014-12-11 project3/paper/paper.tex
more edits through part of Section 4

\item One of $\theta_1, \theta_2, \dots, \theta_n$ is greater than $\pi/2$:

Then, if $\theta_1 > \pi/2$ we see that $\cos^2 \theta_1 = \cos^2 (\pi - \theta_1)$ which is equal to $\cos^2 (\theta_2 + \theta_3 + \ldots + \theta_n)$. Furthermore, $\pi - \theta_1 = \theta_2 + \theta_3 + \ldots + \theta_n < \pi/2$, which means $\cos^2 \theta_1 = \cos^2 (\theta_2 + \theta_3 + \ldots + \theta_n) < \cos^2 \theta_n$. (since $\theta_n < \theta_2 + \theta_3 + \ldots + \theta_n$)

In addition, we see that for all $i \in \{2, 3, \ldots, n-1\}$, $\cos^2 \theta_i < \cos^2 \theta_n$, from which we can conclude that in this case too, $\alpha$ is equal to $\cos^2 \theta_n$.

@@

Since we don't care about the sign of the dot products between any two vectors (we're looking at the squares of the dot products in the above equation), and since $$|v \cdot v'| = |v \cdot (-v')|$$ for some arbitrary vector $v'$, we see that it's easy to transform the three vectors $v_1$, $v_2$ and $v_3$ so that they all lie in the same semi-circle -- to accomplish this, at most one vector needs to be reflected about the origin (multiplied by $-1$).

@@

In the search for a lower bound on $\epsilon$ for arbitrary $(n, N)$ pairs, it is natural to consider the so-called \emph{Gram matrix} of our collection of vectors. 
\begin{definition}
The Gram matrix of a collection of $N$ vectors $v_1,\ldots, v_N$ in $\mathbb R^n$ is the $N$-by-$N$ matrix given by
\[G(v_1,\ldots, v_N)=\left(\begin{matrix}
v_1\cdot v_1 & v_1\cdot v_2 & \ldots & v_1\cdot v_n\\
v_2\cdot v_2 & v_2\cdot v_2 & \ldots & v_2\cdot v_n\\
\vdots & \vdots & \ddots & \vdots\\
v_n\cdot v_1 & v_n\cdot v_2 & \ldots & v_n\cdot v_n\end{matrix}\right).\]
\end{definition}
Note that the Gram matrix can be written as the product $A^TA$, where $A$ is the $n$-by-$N$ matrix
\[A=\left(\begin{matrix}
| & | & \cdots & | & |\\
v_1 & v_2 & \cdots & v_{N-1} & v_N\\
| & | & \cdots & | & | \end{matrix}\right).\]
We can use a well-known lemma on the rank of certain real symmetric matrices to derive a useful lower bound. Our proof of the lemma follows that of a paper by Noga Alon\footnote{Lemma 2.2, \emph{Perturbed identity matrices have high rank: proof and applications}, Noga Alon.}.

@@

Recall the Cauchy-Schwarz for $\mathbb R^n$: given $(x_1,\ldots, x_n)$ and $(y_1,\ldots, y_n)\in \mathbb R^n$, we have
\[\left(\sum_{i=1}^nx_iy_i\right)^2\leq \left(\sum_{i=1}^n x_i^2\right)\left(\sum_{i=1}^ny_i^2\right).\]
In our case, let the $x_i$ equal $\lambda_i$, and let each $y_i$ equal 1 when $\lambda_i$ is nonzero and zero when $\lambda_i$ is zero. We get
\[n^2=\left(\sum_{i=1}^n \lambda_i\right)^2\leq \left(\sum_{i=1}^n \lambda_i^2\right) k\]
so that
\[\sum_{i=1}^n \lambda_i^2\geq\frac{n^2}{k}\geq \frac{n^2}{d}.\]
In fact, the sum $\sum_{i=1}^n \lambda_i^2$ is equal to the trace of $A^TA$, which we can compute explicitly to be $\sum_{i,j}a_{i,j}^2$. Hence we have that
\[\sum_{i,j}a_{i,j}^2\geq \frac{n^2}{d}.\]
But because $|a_{i,j}|\leq \sqrt\epsilon$ for $i\neq j$, we can bound the left hand side:
\[n+n(n-1)\epsilon\geq \left(\sum_{i=1}^j a_{i,i}^2\right)+\left(\sum_{i\neq j}a_{i,j}^2\right)=\sum_{i,j}a_{i,j}^2.\]
We obtain, therefore, that
\[n+n(n-1)\epsilon\geq \frac{n^2}{d},\]
and it clearly follows that
\[\epsilon\geq \frac{n-d}{d(n-1)}.\]
\end{proof}
Using this lemma, we can now prove our lower bound on $\epsilon$.
\begin{theorem}
\label{lower_bound}
For any choice of $n$ and $N$, the minimum value of $\epsilon$ satisfies
\[\epsilon\geq \frac{N-n}{n(N-1)}.\]
\end{theorem}
\begin{proof}
Pick any collection of $N$ unit vectors $v_1,\ldots, v_N$ in $\mathbb R^n$, and define $\epsilon$ as usual. Recall the definitions of $A$ and $G$:
\[A=\left(\begin{matrix}
| & | & \cdots & | & |\\
v_1 & v_2 & \cdots & v_{N-1} & v_N\\
| & | & \cdots & | & | \end{matrix}\right).\] 
\[G(v_1,\ldots, v_N)=A^TA=\left(\begin{matrix}
v_1\cdot v_1 & v_1\cdot v_2 & \ldots & v_1\cdot v_n\\
v_2\cdot v_2 & v_2\cdot v_2 & \ldots & v_2\cdot v_n\\
\vdots & \vdots & \ddots & \vdots\\
v_n\cdot v_1 & v_n\cdot v_2 & \ldots & v_n\cdot v_n\end{matrix}\right).\]
Let $d$ be the rank of $G$. Since $A$ has rank at most $n$, so does $A^T$; since the image space of $G$ must be contained in the image space of $A^T$, we have that $d\leq n$. Note that $G$ is real and symmetric, with $1$s along the diagonal and all other entries with absolute value bounded by $\sqrt \epsilon$. We can apply the lemma to conclude
\[\epsilon\geq \frac{N-d}{d(N-1)}.\]
Since $d\leq n$, we have as desired
\[\epsilon\geq \frac{N-n}{n(N-1)}.\]
\end{proof}
We present now a table of the bounds the theorem gives to us for varying $n$ and $N$:
\begin{table}[h]
   \centering
    \begin{tabular}{ | c | c | c | c |}
    \hline
      & $n=2$ & $n=3$ & $n=4$ \\ \hline
    $N=2$ & $0$ & $-$ & $-$ \\ \hline
    $N=3$ & $1/4$ & $0$ & $-$ \\ \hline
    $N=4$ & $1/3$ & $1/9$ & $0$ \\ \hline
    $N=5$ & $3/8$ & $1/6$ & $1/16$ \\ \hline
    $N=6$ & $2/5$ & $1/5$ & $1/10$ \\ \hline
    $N=8$ &  $3/7$ & $5/21$ & $1/7$ \\ \hline
    $N=12$ & $5/11$ & $3/11$ & $2/11$ \\
    \hline
    \end{tabular}
    \caption {Lower bounds on $\epsilon$ for different values of $(n, N)$, given by Theorem \ref{lower_bound}.}
    \end{table}

@@

\section{Some Optimal Configurations}
Below we provide choices of $n$ and $N$ for which we can achieve the lower bound given by Theorem ~\ref{lower_bound}.
\subsection{$n=3,N=6$}
In this situation we have a lower bound of $\epsilon\ge\frac{1}{5}$ from the above theorem. Consider the following 6 vectors on the unit sphere in $\R^3$ with variables $a$ and $b$. 
$$
(a,b,0),(a,0,b),(b,0,a)
$$
$$
(a,-b,0),(a,0,-b),(-b,0,a)
$$
There are three possible dot products we get from computing dot products between non-equal vectors chosen from these six vectors: $ab,-ab,a^2-b^2$. This gives us two possible values of $\epsilon$: $(a^2-b^2)^2,(ab)^2$. It would seem that an optimal configuration can be obtained by setting $a^2-b^2=ab$. If we add in the constraint that $a^2+b^2=1$ (since the six vectors we're interested in must live on the unit sphere), we get
$$
a=\sqrt{\frac{1}{10}(5-\sqrt{5}}, b=-\frac{1+\sqrt{5}}{2}\sqrt{\frac{1}{10}(5-\sqrt{5}}. 
$$
If we compute $\epsilon=(a^2-b^2)^2=(ab)^2$, we see that $\epsilon=\frac{1}{5}$; coupling this knowledge with the fact that for $(n, N) = (3, 6)$, we see that these six points give us an optimal configuration.
\subsection{$N=n+1$}
Let us now consider the situation in which we try to minimize $\epsilon$ for $n+1$ vectors in an $n$ dimensional space. From Theorem ~\ref{lower_bound} we know that $\epsilon\ge \frac{1}{n^2}$, so if we show that there exist some $n+1$ vectors in $n$-space that achieve an $\epsilon$ equal to $1/n^2$, then we can conclude that $1/n^2$ is the optimal value of $\epsilon$ for $N = n+1$.

To do this, we first define what we mean by an $n$-dimensional simplex.

@@

\begin{theorem}
If $N=n+1$ and the vectors in question are denoted $v_1,\cdots, v_N$, then over all configurations of these vectors the minimum value of $\epsilon$ for $i\neq j$ is $\frac{1}{^2n}$. This is achieved when the vectors are arranged in an $n$ dimensional simplex centered at the origin.
\label{simplex}
\end{theorem}

@@

We must first describe the vectors that make up the origin centered simplex. To center our simplex at the origin we take each vector in the original simplex, and subtract from it the vector representing the centroid of the original simplex. The centroid of the original simplex takes the form $(\frac{1}{N},\frac{1}{N},\ldots,\frac{1}{N})$, so the points of the new simplex take the form
$$
\left(-\frac{1}{N},\cdots,-\frac{1}{N},\frac{N-1}{N},-\frac{1}{N},\cdots,-\frac{1}{N}\right).
$$
If we renormalize these vectors so that each vector has unit length, we get 
$$
\sqrt{\frac{N}{N-1}}\left(-\frac{1}{N},\cdots,-\frac{1}{N},\frac{N-1}{N},-\frac{1}{N},\cdots,-\frac{1}{N}\right).
$$
Now, taking the dot product between any two such vectors gives us
$$
\frac{N}{N-1}\left(-2\cdot\frac{N-1}{N^2}+(N-2)\cdot\frac{1}{N^2}\right)=\frac{N}{N-1}\cdot-\frac{1}{N}=-\frac{1}{N-1}=-\frac{1}{n}.
$$
So for this configuration $\epsilon=\frac{1}{n^2}$. By Theorem ~\ref{lower_bound} we know that $\epsilon\ge \frac{1}{n^2}$, so we have achieved the lower bound. 
\end{proof}
\section{Numerical analysis and associated conjectures}

>>>>

\item One of $\theta_1, \theta_2, \dots, \theta_n$ is greater than $\pi/2$: Then in particular $\theta_1 > \pi/2$. We see that $\cos^2 \theta_1 = \cos^2 (\pi - \theta_1)$ which is equal to $\cos^2 (\theta_2 + \theta_3 + \ldots + \theta_n)$. Furthermore, $\pi - \theta_1 = \theta_2 + \theta_3 + \ldots + \theta_n < \pi/2$, which means $\cos^2 \theta_1 = \cos^2 (\theta_2 + \theta_3 + \ldots + \theta_n) < \cos^2 \theta_n$. (since $\theta_n < \theta_2 + \theta_3 + \ldots + \theta_n$)

In addition, we see that for all $i \in \{2, 3, \ldots, n-1\}$, $\cos^2 \theta_i < \cos^2 \theta_n$, from which we can conclude that $\alpha$ is still equal to $\cos^2 \theta_n$.

@@

Because the sign of the dot product of any two vectors does not matter, and because $$|v \cdot v'| = |v \cdot (-v')|$$ for some arbitrary vector $v'$, we see that it's easy to transform the three vectors $v_1$, $v_2$ and $v_3$ so that they all lie in the same semi-circle -- to accomplish this, at most one vector needs to be reflected about the origin (multiplied by $-1$).

@@

In the search for a lower bound on $\epsilon$ for arbitrary $(n, N)$ pairs, it is natural to consider the so-called \emph{Gram matrix} of our collection of vectors. 
\begin{definition}
The Gram matrix of a collection of $N$ vectors $v_1,\ldots, v_N$ in $\mathbb R^n$ is the $N$-by-$N$ matrix given by
\[G(v_1,\ldots, v_N)=\left(\begin{matrix}
v_1\cdot v_1 & v_1\cdot v_2 & \ldots & v_1\cdot v_n\\
v_2\cdot v_2 & v_2\cdot v_2 & \ldots & v_2\cdot v_n\\
\vdots & \vdots & \ddots & \vdots\\
v_n\cdot v_1 & v_n\cdot v_2 & \ldots & v_n\cdot v_n\end{matrix}\right).\]
\end{definition}
Note that the Gram matrix can be written as the product $A^TA$, where $A$ is the $n$-by-$N$ matrix
\[A=\left(\begin{matrix}
| & | & \cdots & | & |\\
v_1 & v_2 & \cdots & v_{N-1} & v_N\\
| & | & \cdots & | & | \end{matrix}\right).\]
We can use a well-known lemma on the rank of certain real symmetric matrices to derive a useful lower bound on $\epsilon$. Our proof of the lemma follows that of a paper by Noga Alon\footnote{Lemma 2.2, \emph{Perturbed identity matrices have high rank: proof and applications}, Noga Alon.}.

@@

Recall the Cauchy-Schwarz for $\mathbb R^n$: given $(x_1,\ldots, x_n)$ and $(y_1,\ldots, y_n)\in \mathbb R^n$, we have
\[\left(\sum_{i=1}^nx_iy_i\right)^2\leq \left(\sum_{i=1}^n x_i^2\right)\left(\sum_{i=1}^ny_i^2\right).\]
In our case, let the $x_i$ equal $\lambda_i$, and let each $y_i$ equal 1 when $\lambda_i$ is nonzero and zero when $\lambda_i$ is zero. We get
\[n^2=\left(\sum_{i=1}^n \lambda_i\right)^2\leq \left(\sum_{i=1}^n \lambda_i^2\right) k\]
so that
\[\sum_{i=1}^n \lambda_i^2\geq\frac{n^2}{k}\geq \frac{n^2}{d}.\]
In fact, the sum $\sum_{i=1}^n \lambda_i^2$ is equal to the trace of $A^TA$, which we can compute explicitly to be $\sum_{i,j}a_{i,j}^2$. Hence we have that
\[\sum_{i,j}a_{i,j}^2\geq \frac{n^2}{d}.\]
But because $|a_{i,j}|\leq \sqrt\epsilon$ for $i\neq j$, we can bound the left hand side:
\[n+n(n-1)\epsilon\geq \left(\sum_{i=1}^j a_{i,i}^2\right)+\left(\sum_{i\neq j}a_{i,j}^2\right)=\sum_{i,j}a_{i,j}^2.\]
We obtain, therefore, that
\[n+n(n-1)\epsilon\geq \frac{n^2}{d},\]
and it clearly follows that
\[\epsilon\geq \frac{n-d}{d(n-1)}.\]
\end{proof}
Using this lemma, we can now prove our lower bound on $\epsilon$.
\begin{theorem}
\label{lower_bound}
For any choice of $n$ and $N$, the minimum value of $\epsilon$ satisfies
\[\epsilon\geq \frac{N-n}{n(N-1)}.\]
\end{theorem}
\begin{proof}
Pick any collection of $N$ unit vectors $v_1,\ldots, v_N$ in $\mathbb R^n$, and define $\epsilon$ as usual. Recall the definitions of $A$ and $G$:
\[A=\left(\begin{matrix}
| & | & \cdots & | & |\\
v_1 & v_2 & \cdots & v_{N-1} & v_N\\
| & | & \cdots & | & | \end{matrix}\right).\] 
\[G(v_1,\ldots, v_N)=A^TA=\left(\begin{matrix}
v_1\cdot v_1 & v_1\cdot v_2 & \ldots & v_1\cdot v_n\\
v_2\cdot v_2 & v_2\cdot v_2 & \ldots & v_2\cdot v_n\\
\vdots & \vdots & \ddots & \vdots\\
v_n\cdot v_1 & v_n\cdot v_2 & \ldots & v_n\cdot v_n\end{matrix}\right).\]
Let $d$ be the rank of $G$. Since $A$ has rank at most $n$, so does $A^T$; since the image space of $G$ must be contained in the image space of $A^T$, we have that $d\leq n$. Note that $G$ is real and symmetric, with $1$s along the diagonal and all other entries with absolute value bounded by $\sqrt \epsilon$. We can apply the lemma to conclude
\[\epsilon\geq \frac{N-d}{d(N-1)}.\]
Since $d\leq n$, we have as desired
\[\epsilon\geq \frac{N-n}{n(N-1)}.\]
\end{proof}
We present now a table of the bounds the theorem gives us for varying $n$ and $N$:
\begin{table}[h]
   \centering
    \begin{tabular}{ | c | c | c | c |}
    \hline
      & $n=2$ & $n=3$ & $n=4$ \\ \hline
    $N=2$ & $0$ & $-$ & $-$ \\ \hline
    $N=3$ & $1/4$ & $0$ & $-$ \\ \hline
    $N=4$ & $1/3$ & $1/9$ & $0$ \\ \hline
    $N=5$ & $3/8$ & $1/6$ & $1/16$ \\ \hline
    $N=6$ & $2/5$ & $1/5$ & $1/10$ \\ \hline
    $N=8$ &  $3/7$ & $5/21$ & $1/7$ \\ \hline
    $N=12$ & $5/11$ & $3/11$ & $2/11$ \\
    \hline
    \end{tabular}
    \caption {Lower bounds on $\epsilon$ for different values of $(n, N)$, given by Theorem \ref{lower_bound}.}
    \end{table}

@@

\section{Some Optimal Configurations}
Below we provide choices of $n$ and $N$ for which we can achieve the lower bound given by Theorem ~\ref{lower_bound}.
\subsection{$n=3,N=6$}
Let us consider collections of six vectors in $\mathbb R^3$. As can be seen in Table 1, the lower bound provided in this case by Theorem \ref{lower_bound} is $\frac 1 5$ as well. We shall now see, in fact, that this lower bound is attainable.
\begin{theorem}
For $n=3$ and $N=6$, the optimum value of $\epsilon$ is given by $\frac 1 5$.
\end{theorem}
\begin{proof}Consider the following 6 vectors on the unit sphere in $\R^3$ with variables $a$ and $b$. 
$$
(a,b,0),(a,0,b),(b,0,a)
$$
$$
(a,-b,0),(a,0,-b),(-b,0,a)
$$
There are three possible dot products we get from computing dot products between non-equal vectors chosen from these six vectors: $ab,-ab,a^2-b^2$. This gives us two possible values of $\epsilon$: $(a^2-b^2)^2,(ab)^2$. It would seem that an optimal configuration can be obtained by setting $a^2-b^2=ab$. If we add in the constraint that $a^2+b^2=1$ (since the six vectors we're interested in must live on the unit sphere), we get
$$
a=\sqrt{\frac{1}{10}(5-\sqrt{5}}, b=-\frac{1+\sqrt{5}}{2}\sqrt{\frac{1}{10}(5-\sqrt{5}}. 
$$
If we compute $\epsilon=(a^2-b^2)^2=(ab)^2$, we see that $\epsilon=\frac{1}{5}$. Coupled with our knowledge that $\frac 1 5$ is a lower bound on $\epsilon$, we conclude that these six points give us an optimal configuration.\end{proof}
\subsection{$N=n+1$}
Let us now consider the situation in which we try to minimize $\epsilon$ for $n+1$ vectors in an $n$ dimensional space. From Theorem \ref{lower_bound} we know that $\epsilon\ge \frac{1}{n^2}$, so it suffices to show there exist some $n+1$ vectors in $n$-space that achieve an $\epsilon$ equal to $1/n^2$.

To do this, we first introduce the notion of an $n$-dimensional simplex.

@@

\begin{theorem}
If $N=n+1$, then the minimum value of $\epsilon$ for $i\neq j$ is $\frac{1}{n^2}$. This $\epsilon$ is achieved when the vectors are arranged in an $n$ dimensional simplex centered at the origin.
\label{simplex}
\end{theorem}

@@

We must first describe the vectors that make up the origin centered simplex. To center our simplex at the origin we take each vector in the original simplex, and subtract from it the vector representing the centroid of the original simplex. The centroid of the original simplex takes the form $(\frac{1}{N},\frac{1}{N},\ldots,\frac{1}{N})$, so the points of the new simplex take the form
$$
\left(-\frac{1}{N},\cdots,-\frac{1}{N},\frac{N-1}{N},-\frac{1}{N},\cdots,-\frac{1}{N}\right).
$$
If we renormalize these vectors so that each vector has unit length, we get 
$$
\sqrt{\frac{N}{N-1}}\left(-\frac{1}{N},\cdots,-\frac{1}{N},\frac{N-1}{N},-\frac{1}{N},\cdots,-\frac{1}{N}\right).
$$
Now, taking the dot product between any two such vectors gives us
$$
\frac{N}{N-1}\left(-2\cdot\frac{N-1}{N^2}+(N-2)\cdot\frac{1}{N^2}\right)=\frac{N}{N-1}\cdot-\frac{1}{N}=-\frac{1}{N-1}=-\frac{1}{n}.
$$
So for this configuration $\epsilon=\frac{1}{n^2}$. By Theorem \ref{lower_bound} we know that $\epsilon\ge \frac{1}{n^2}$, so we have achieved the lower bound. 
\end{proof}
\section{Numerical analysis and associated conjectures}
